---
title: "glmnetUI Model Report"
format:
  html:
    toc: true
    theme: flatly
    mainfont: "Roboto Condensed"
    html-math-method: katex
    embed-resources: true
    include-in-header:
      text: |
        <style>
        @import url('https://fonts.googleapis.com/css2?family=Roboto+Condensed:wght@400;700&display=swap');
        body { font-family: 'Roboto Condensed', sans-serif; }
        </style>
  pdf:
    toc: true
    mainfont: "Roboto Condensed"
    sansfont: "Roboto Condensed"
    include-in-header:
      text: |
        \usepackage{unicode-math}
        \setmathfont{Latin Modern Math}
        \setmathfont{Roboto Condensed}[range=up/{num,latin,Latin}]
        \setmathfont{Roboto Condensed}[range=it/{latin,Latin}]
        \usepackage{longtable}
        \usepackage{booktabs}
        \usepackage{lscape}
  docx:
    toc: true
    reference-doc: reference.docx
params:
  data_file: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

if (nzchar(params$data_file) && file.exists(params$data_file)) {
  .assets <- readRDS(params$data_file)
  .plots_dir <- file.path(dirname(params$data_file), "plots")
  .has_assets <- dir.exists(.plots_dir) &&
    length(list.files(.plots_dir)) > 0L
} else {
  stop("No data_file provided.")
}

fmt <- if (!is.null(knitr::pandoc_to())) knitr::pandoc_to() else "html"
.plot_ext <- if (fmt == "latex") "pdf" else "png"

include_plot <- function(name, width = 8, height = 5) {
  if (.has_assets) {
    f <- file.path(.plots_dir, paste0(name, ".", .plot_ext))
    if (file.exists(f)) {
      # Use relative path — plots/ is next to the QMD in the render dir.
      # Absolute paths break in LaTeX (luatex prepends ./).
      rel <- file.path("plots", paste0(name, ".", .plot_ext))
      cat(sprintf("![](%s)\n\n", rel))
    }
  }
}
```

## Dataset Description

| Property | Value |
|----------|-------|
| Data file | `r .assets$data_file_name` |
| Observations | `r .assets$n_obs` |
| Response variable | `r .assets$response` |
| Predictors | `r paste(.assets$predictors, collapse = ", ")` |
| Purpose | `r .assets$purpose` |

## Model Specification

| Parameter | Value |
|-----------|-------|
| Alpha | `r .assets$alpha` |
| Lambda | `r signif(.assets$lambda, 5)` |
| Lambda selection | `r .assets$lambda_method` |
| Family | `r .assets$family` |
| Standardize | `r .assets$standardize` |
| Relaxed | `r .assets$relaxed` |
| Non-zero coefficients | `r .assets$n_nonzero` |
| Total predictors in model | `r .assets$n_predictors` |

## Results Summary

| Metric | Value |
|--------|-------|
| R² | `r sprintf("%.4f", .assets$r_squared)` |
| Adjusted R² | `r sprintf("%.4f", .assets$adj_r_squared)` |
| Generalized R² | `r sprintf("%.4f", .assets$gen_r_squared)` |
| CV R² | `r if (!is.na(.assets$cv_r_squared)) sprintf("%.4f", .assets$cv_r_squared) else "N/A"` |
| RMSE | `r sprintf("%.2f", .assets$rmse)` |
| MAE | `r sprintf("%.2f", .assets$mae)` |

## Model Equation

```{r equation, results='asis'}
# `\small` is a text-mode command, invalid inside `$$...$$` math; it breaks
# pandoc's LaTeX -> Word (OMML) conversion (raw LaTeX in the doc). Strip for
# Word only. Emit as a standalone block (blank lines around `$$`) via cat so
# the multi-line equation is parsed as display math in every format.
.eq <- .assets$equation_latex
if (identical(fmt, "docx")) {
  .eq <- gsub("\\\\small", "", .eq)
  # Drop any now-blank lines: a blank line inside `$$...$$` terminates the math
  # block for pandoc and breaks the Word (OMML) conversion.
  .lines <- strsplit(.eq, "\n", fixed = TRUE)[[1]]
  .eq <- paste(.lines[nzchar(trimws(.lines))], collapse = "\n")
}
cat("\n$$\n", .eq, "\n$$\n\n", sep = "")
```

## Coefficients

```{r coefficients}
knitr::kable(.assets$coef_df, digits = 4)
```

## Variable Importance

```{r importance-plot, results='asis'}
include_plot("importance")
```

```{r importance-table}
knitr::kable(.assets$importance_df, digits = 4)
```

## Contributions

```{r contributions, results='asis'}
contrib_names <- .assets$contrib_names
for (vname in contrib_names) {
  safe_name <- gsub("[^a-zA-Z0-9_]", "_", vname)
  # Use a markdown heading with the variable name as inline code. Pandoc then
  # escapes underscores/specials correctly for every format. Emitting raw
  # LaTeX (\subsection{living_sqft}) breaks PDF with "Missing $ inserted".
  cat(sprintf("\n### `%s`\n\n", vname))
  is_int <- grepl(":", vname, fixed = TRUE)
  if (is_int) {
    # Interaction: scatter + heatmap + 3D persp
    include_plot(paste0("contrib_", safe_name, "_scatter"))
    include_plot(paste0("contrib_", safe_name, "_heatmap"))
    include_plot(paste0("contrib_", safe_name, "_3d"))
  } else {
    include_plot(paste0("contrib_", safe_name))
  }
  cat("\n\n")
}
```

## Correlation Matrix

```{r correlation, results='asis'}
include_plot("correlation")
```

## ANOVA Decomposition

```{r anova}
knitr::kable(.assets$anova_df, digits = 2)
```

## Diagnostics

### Coefficient Path

```{r coef-path, results='asis'}
include_plot("coef_path")
```

### Cross-Validation Error

```{r cv-error, results='asis'}
include_plot("cv_error")
```

### Actual vs Predicted

```{r actual-vs-predicted, results='asis'}
include_plot("actual_vs_predicted")
```

### Residuals vs Fitted

```{r residuals-vs-fitted, results='asis'}
include_plot("residuals_vs_fitted")
```

### Normal Q-Q Plot

```{r qq-plot, results='asis'}
include_plot("qq")
```

## Model Output

```{r model-output, comment=""}
cat("== glmnet Model ==\n\n")
print(.assets$model)
```

---

*Report generated by [glmnetUI](https://github.com/wcraytor/glmnetUI) on `r Sys.time()`*
