---
title: "Getting Started with glmnetUI for Real Estate Appraisers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with glmnetUI for Real Estate Appraisers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is glmnetUI?

glmnetUI is an interactive Shiny application that makes elastic net regression
accessible through a point-and-click interface. It wraps the powerful `glmnet`
R package in a user-friendly application designed for real estate appraisers
and general regression users.

**Elastic net regression** combines the strengths of two regularization
techniques:

- **Ridge regression** (alpha = 0): Shrinks coefficients toward zero but keeps
  all variables in the model.
- **Lasso regression** (alpha = 1): Can shrink coefficients exactly to zero,
  effectively performing variable selection.
- **Elastic net** (0 < alpha < 1): A blend of both approaches.

## Launching the Application

```{r launch}
library(glmnetUI)
glmnetUI()
```

This opens the application in your default web browser.

## Step 1: Import Your Data

1. Click **"Upload CSV or Excel File"** in the sidebar.
2. Select a CSV (`.csv`) or Excel (`.xls`, `.xlsx`) file from your computer.
3. For Excel files, choose the appropriate sheet from the dropdown.
4. A preview of your data appears in the table below.

### Supported Formats

| Format | Extension | Reader |
|--------|-----------|--------|
| CSV | .csv | readr |
| Excel | .xls, .xlsx | readxl |

## Step 2: Assign Variable Roles

After importing data, assign each column a role:

- **Response Variable**: The value you want to predict (e.g., sale price).
- **Predictor Variables**: The features used to make predictions
  (e.g., square footage, lot size, age).
- **Ignored Columns**: Columns not used in modeling (e.g., address, parcel ID).

### Expected Coefficient Signs

For numeric predictors, you can specify whether you expect the coefficient to
be **positive**, **negative**, or **either**. This is critical for appraisers:

- **Square footage**: Expected **positive** (more space = higher value).
- **Age**: Expected **negative** (older properties typically worth less).
- **Lot size**: Expected **positive**.

If the model produces a coefficient with an unexpected sign, it will be
**flagged in red** in the coefficient table. This helps you identify and
defend potentially problematic model results.

## Step 3: Configure the Model

### Alpha (Mixing Parameter)

The alpha slider controls the type of regularization:

- **0 (Ridge)**: Keeps all variables, shrinks coefficients proportionally.
  Good when you believe all variables are relevant.
- **1 (Lasso)**: Performs variable selection by setting some coefficients to
  exactly zero. Good for identifying the most important predictors.
- **Between 0 and 1**: A blend. Start with the default of 1 (Lasso) for
  appraisal work, as it produces simpler, more defensible models.

### Lambda Selection

Lambda controls the strength of regularization:

- **Cross-validation (automatic)**: Recommended. The application tests many
  lambda values and selects the best one using 10-fold cross-validation.
  - **lambda.1se**: More regularized (simpler model). Default and recommended
    for appraisal.
  - **lambda.min**: Least regularized that minimizes prediction error.
- **Manual**: Enter a specific lambda value if you have a reason to.

### Family

Choose the distribution family for your response variable:

- **gaussian**: For continuous responses (e.g., sale price). Most common.
- **binomial**: For binary outcomes (e.g., sold/not sold).
- **poisson**: For count data.

### Standardize

Keep this checked (default). Standardizing ensures all predictors are on the
same scale before fitting, which is important for regularization to work
correctly.

## Step 4: Fit the Model

Click **"Fit Model"** to run the analysis. The application will:

1. Prepare the data (handle missing values, encode factors).
2. Run cross-validation (if selected) to find the optimal lambda.
3. Fit the final model.
4. Display results in the main panel tabs.

## Step 5: Review Results

### Coefficients Tab

Shows a table of all non-zero coefficients with:

- **Variable name**: The predictor.
- **Coefficient value**: The estimated effect.
- **Sign Warning**: Flags coefficients with unexpected signs in red.

Review sign warnings carefully. An unexpected sign may indicate:

- Multicollinearity among predictors.
- A variable that needs to be removed or transformed.
- An unexpected but genuine market relationship.

### Diagnostics Tab

Four diagnostic plots help you evaluate model quality:

1. **Coefficient Path**: Shows how coefficients change as lambda increases.
   Helps visualize variable selection.
2. **CV Error**: Shows cross-validation error across lambda values.
   The dashed lines mark lambda.min and lambda.1se.
3. **Actual vs Predicted**: Points should cluster around the 45-degree line.
   Scatter indicates prediction error.
4. **Residuals vs Fitted**: Should show random scatter around zero.
   Patterns suggest model misspecification.

Each plot has a **Download PNG** button for saving.

## Step 6: Export a Report

Fill in the report fields:

- **Appraiser Name**: Your name for the report header.
- **Property Address**: The subject property.
- **Report Date**: Defaults to today.
- **File Number**: Your internal reference.

Then click:

- **Export to Word (.docx)**: Creates a formatted Word document with all
  tables and plots.
- **Export to PDF**: Creates a PDF report via rmarkdown.

Reports include the coefficient table with sign warnings highlighted and
all four diagnostic plots.

## Tips for Appraisers

1. **Start with Lasso (alpha = 1)** to identify the most important variables.
2. **Use lambda.1se** for simpler, more defensible models.
3. **Set expected signs** for all numeric predictors before fitting.
4. **Review sign warnings** carefully and be prepared to explain any
   unexpected signs in your appraisal report.
5. **Use the Actual vs Predicted plot** to assess overall model accuracy.
6. **Export reports** to include in your work file for documentation.

## Example Workflow

```{r example, eval = FALSE}
# Launch the application
library(glmnetUI)
glmnetUI()

# In the app:
# 1. Upload your sales data CSV
# 2. Set sale_price as response
# 3. Select sqft, lot_size, age, bedrooms, bathrooms as predictors
# 4. Set expected signs: sqft=positive, lot_size=positive,
#    age=negative, bedrooms=positive, bathrooms=positive
# 5. Keep alpha=1 (Lasso), CV with lambda.1se
# 6. Click "Fit Model"
# 7. Review coefficients and sign warnings
# 8. Export Word report for your work file
```
