---
title: "Getting Started with glmnetUI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with glmnetUI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

`glmnetUI` provides an interactive Shiny interface for elastic net regression
modeling using the `glmnet` package. It runs as a local Shiny application ---
no login, no server, no accounts. You launch it from R, import a dataset
(CSV or Excel), configure your model, and fit it interactively.

**Elastic net regression** combines two regularization techniques:

- **Ridge regression** (alpha = 0): Shrinks coefficients toward zero but keeps
  all variables.
- **Lasso regression** (alpha = 1): Sets some coefficients exactly to zero,
  performing automatic variable selection.
- **Elastic net** (0 < alpha < 1): A blend of both approaches.

### Three Purpose Modes

A **Purpose** radio button at the top of the sidebar selects:

- **General** --- Elastic net regression for any dataset. The default mode.
  A **Skip first row** checkbox lets you exclude row 1 from fitting.
- **For Appraisal** --- Tailored for real estate appraisal with subject
  property handling, special column designations, RCA adjustments, and
  formatted report export.
- **Market Area Analysis** --- For analyzing groups of properties with optional
  first row exclusion.

Settings are saved per file and purpose combination, so switching modes
preserves each mode's configuration independently.

### Launching the Application

```{r}
library(glmnetUI)
glmnetUI()
```

This opens the application at `http://localhost:7879`.

## Basic Workflow

### 1. Import Data

Click **"Upload CSV or Excel File"** in the sidebar. Supported formats:

| Format | Extension | Notes |
|--------|-----------|-------|
| CSV | .csv | Locale-aware separator and decimal mark |
| Excel | .xls, .xlsx | Sheet selector for multi-sheet workbooks |

Column names are automatically converted to `snake_case` on import.

### 2. Configure Variables

After importing, a predictor table shows each column with:

- **Type** --- data type dropdown (numeric, integer, character, Date, POSIXct)
- **Fac** --- factor checkbox (treat as categorical)
- **Inc** --- checkbox to include as predictor
- **Force** --- force into model (penalty factor = 0, never dropped by lasso)
- **Sign** --- expected coefficient sign (positive, negative, or either)
- **NAs** --- missing value count
- **Special** --- special column designation (appraisal/market modes only)

Select your **response variable** from the dropdown, then check/uncheck
predictors as needed.

### 3. Set Model Parameters

Under **"4. glmnet Call Parameters"**:

- **Alpha**: 0 = Ridge, 1 = Lasso (default), or use grid search
- **Lambda**: Cross-validation (recommended) or manual entry
- **Lambda choice**: `lambda.1se` (simpler model) or `lambda.min` (best fit)
- **Family**: gaussian (continuous), binomial (binary), or poisson (counts)
- **Standardize**: Recommended on (default)
- **Enforce sign constraints**: Force coefficients to match expected signs
- **Relaxed lasso**: Refit surviving variables with less shrinkage
- **Interaction matrix**: Check pairs to allow interaction terms
  (Warning: glmnet interactions produce adjustments that can be hard to
  interpret in court or audit settings.)

### 4. Fit the Model

Click **"Fit Model"**. The application:

1. Prepares data (missing values, factor encoding, weights)
2. Runs cross-validation to find optimal lambda
3. Fits the final model
4. Populates all result tabs

### 5. Review Results

Results appear in the main panel tabs:

- **Data Preview** --- Raw data table
- **Equation** --- LaTeX-rendered model equation
- **Correlation** --- Heatmap of predictor correlations (no model needed)
- **Summary** --- R², GR², CV R², RMSE, MAE; appraisal metrics (COD, PRD)
- **Coefficients** --- Non-zero coefficients with sign warnings
- **Variable Importance** --- Bar chart of |coef| x sd(x) per predictor
- **Contributions** --- Per-variable effect plots: scatter with fit line
  (numeric), box plot (factor), scatter/heatmap/3D surface (interactions)
- **ANOVA** --- Variance decomposition table
- **Diagnostics** --- Coefficient path, CV error, actual vs predicted, residuals
- **Report** --- Configure and export reports

### 6. Export

- **Download Output** (all modes): Excel with predictions, residuals, CQA
  scores, and per-variable contributions
- **Sales Grid** (appraisal mode): Generate a formatted Sales Comparison
  Grid Excel workbook from RCA output. A modal dialog lets you select
  comps (recommended comps with gross adjustment < 25% are pre-checked).
  Up to 30 comps across 10 sheets (3 per sheet) with grouped rows,
  formulas, and unlocked residual cells for appraiser input.
- **Download Report**: Word (.docx), PDF, or HTML with all tables and plots.
  Reports include contribution plots matching the Contributions tab
  (scatter, heatmap, 3D persp snapshots for interactions). Rendered via
  Quarto with automatic rmarkdown fallback.

## Tips

1. **Start with Lasso (alpha = 1)** to identify important variables.
2. **Use lambda.1se** for simpler, more defensible models.
3. **Set expected signs** before fitting and review warnings after.
4. **Use interactions cautiously** -- glmnet interactions can produce adjustments
   that are hard to interpret for court or audit purposes.
5. **Try relaxed lasso** if coefficients appear over-shrunk.

## Demo Dataset

glmnetUI includes the `Appraisal_1.csv` demo dataset:

```{r}
demo_file <- system.file("extdata", "Appraisal_1.csv", package = "glmnetUI")
```

Or import through the app's file upload.

See the **User Guide** vignette for complete documentation of all features,
special column types, RCA calculations, and the demo dataset.
