---
title: "glmnetUI User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{glmnetUI User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`glmnetUI` is a graphical user interface for the R `glmnet` package, which
implements elastic net regularized regression. It runs as a local Shiny
application --- no login, no server, no accounts. You launch it from R,
import a dataset (CSV or Excel), configure your model, and fit it
interactively.

### Three Purpose Modes

When you launch glmnetUI, a **Purpose** radio button at the top of the sidebar
lets you choose one of three modes:

- **General** --- Elastic net regression for any dataset. The default mode.
- **For Appraisal** --- Tailored for real estate appraisal with subject
  property handling, special column designations, RCA adjustments, and
  formatted report export.
- **Market Area Analysis** --- For analyzing groups of properties with optional
  subject row exclusion.

### Getting Started

```{r eval=FALSE}
library(glmnetUI)
glmnetUI()
```

This opens the Shiny app at `http://localhost:7879`.

## Locale & Regional Settings

glmnetUI supports international number, date, and CSV formatting conventions
through a country-based locale system. The **Settings** dropdown in the title
bar contains **Country** and **Paper** selectors for 31 supported countries.
Each preset configures:

- **CSV separator** --- comma (`,`) for US/UK/Japan or semicolon (`;`) for
  most of Europe, where the comma is used as a decimal mark.
- **Decimal mark** --- period (`.`) or comma (`,`).
- **Thousands separator** --- comma (US/UK/Japan), period (Germany/Italy/Spain),
  space (Finland/France/Poland/Baltics/Ukraine/Russia), or apostrophe
  (Switzerland).
- **Paper size** --- Letter (US/Canada/Mexico) or A4 (everywhere else).

### Saving Defaults

Click **Save as my default** to store your locale preferences globally. These
defaults apply to all future sessions regardless of which data file you load.

## MLS Input Data Requirements

For real estate workflows, your input data typically comes from an MLS export.
glmnetUI accepts CSV and Excel files. On import, column names are automatically
converted to `snake_case`, e.g. "Area ID" gets converted to "area_id". The CSV
separator and decimal mark used during import are determined by the locale
settings.

### Required Columns for Appraisal Mode

The full appraisal workflow benefits from these columns:

| Column | Special Type | Purpose |
|:-------|:-------------|:--------|
| Sale Price (or equivalent) | (target) | Response variable |
| Contract Date | `contract_date` | Computes `sale_age` from effective date |
| Listing Date | `listing_date` | Fallback DOM = contract date - listing date |
| Days on Market | `dom` | Displayed in exports |
| Concessions | `concessions` | Net SP = Sale Price - Concessions |
| Living Area (SF) | `living_area` | Per-SF residuals (`residual_sf`, `cqa_sf`) |
| Lot Size | `lot_size` | Grouped in "Site Size" row |
| Site Dimensions | `site_dimensions` | Grouped with lot size |
| Latitude | `latitude` | Rounded to 3 dp; Location group |
| Longitude | `longitude` | Rounded to 3 dp; Location group |
| Area ID | `area` | Market area/neighborhood identifier |
| Actual Age | `actual_age` | Grouped in "Age" row |
| Effective Age | `effective_age` | Grouped with actual age |
| Address | `display_only` | Shown in exports; excluded from model |

Spreadsheet column names can be in any language. The "special" names are
in English so that the R program can give them special treatment. Otherwise,
the given column names show up in the regression models, graphs and, if doing
appraisals, the output reports.

Not all columns are required. glmnetUI adapts --- if a column is missing, the
corresponding feature is simply omitted.

### Key Predictors for Real Estate Models

For real estate pricing models, certain columns are highly recommended:

1. **Sale Age** --- days between the contract sale date and the effective date
   of the appraisal. If multi-year sales history is being used, sale_age often
   plays a central role in estimating the sale price.

2. **Living Area** --- also called "Living Sqft," "GLA" (gross living area),
   etc. A leading determinant of sale price.

3. **Total Bath Count** --- total number of full, quarter, half, and 3/4
   bathrooms. For example, two full baths and one half-bath = 2.5.

4. **Garage Bays** or **Garage Area** --- number of garage spaces or
   garage square footage.

5. **Lot Size** --- land area of the property, typically in square feet or
   acres.

6. **Longitude**, **Latitude**, and **Area ID** --- location variables.

### Data Quality

- **Missing values**: Rows with NAs in predictors or target are removed before
  fitting.
- **Date columns**: Character columns matching common date formats are
  auto-parsed to Date/POSIXct on import.
- **Numeric columns**: Sale price, living area, etc. must be numeric (no
  dollar signs or commas).

### Subject Row Placement

In **Appraisal** mode, row 1 must be the subject property. In **Market**
mode, placing the subject in row 1 is optional. In **General** mode, all
rows are treated equally.

## Projects

glmnetUI organizes work into **projects** under a shared `regProj` root folder
(the same tree used by earthUI and mgcvUI). Each project owns an input folder
(`<os>_in`) and per-engine output folders (e.g. `<os>_out_glmnet`).

### Creating a project

In **Section 1 (Project)**, click **+ New…** and provide:

- **Purpose** --- General, Appraisal, or Market Area.
- **Location** --- the project's political/administrative hierarchy. The set of
  levels you fill in depends on the **country** you choose, so foreign naming
  conventions are supported automatically (see below).
- **Project Name** --- a short label, **limited to 8 characters**
  (letters, digits, `_` and `-`).

### Automatic folder naming

When you create a project, the app does **not** use your typed name verbatim.
It builds a unique, time-stamped folder name by prepending the location codes
and the **creation date-time** in front of your name:

```
<country>_<admin level codes…>_<YYYYMMDD-HHMMSS>_<your name>
```

For example, a U.S. project named `burl` created on 4 June 2026 at 13:15:00
becomes:

```
us_ca_081_burlin_20260604-131500_burl
```

where `YYYYMMDD-HHMMSS` is the actual current date and time at the moment of
creation. The 8-character limit applies only to the part **you** type; the
country, admin-level, and timestamp segments are added for you.

### Foreign political naming conventions

The admin-level segments follow **each country's own hierarchy**, not a fixed
state/county/city scheme. The number and meaning of the levels vary by country,
for example:

| Country     | Admin levels (top → bottom)                |
|-------------|--------------------------------------------|
| US          | state / county / city                      |
| Great Britain | region / district / place                |
| France      | region / département / commune             |
| Germany     | bundesland / kreis / gemeinde              |
| Japan       | prefecture / city-or-ward / district       |
| Singapore   | planning area (single level)               |
| *(other)*   | region / city (generic fallback)           |

The timestamp-and-name portion is independent of the number of levels, so the
naming works for any of these conventions. (For a U.S. project the prefix is
`us_<state>_<county>_<city>_`; for a French one it is
`fr_<region>_<département>_<commune>_`; for Singapore it is `sg_<area>_`.)

**Existing projects are unaffected** --- only newly created projects get the
timestamp prefix; projects already on disk keep their original names and remain
fully accessible.

## The Sidebar Workflow

The sidebar is organized into numbered sections:

1. **Import Data** --- CSV or Excel file upload
2. **Import from earthUI (optional)** --- Import an earthUI result `.rds`
   file to use earth's hinge basis functions with glmnet's regularization.
   The browse button is styled to match the Section 1 file upload.
3. **Project Output Folder** --- where downloads are saved
4. **Variable Configuration** --- target, predictors, data types, expected
   signs. Special column designations appear in appraisal/market only.
   Effective date input appears in appraisal/market only.
5. **glmnet Call Parameters** --- alpha, lambda, family, standardize, sign
   constraints, relaxed lasso, interaction matrix, advanced parameters
6. **Fit Glmnet Model** --- button to run the model
7. **Download Output** --- Excel file with predictions, residuals, CQA
   scores, and per-variable contributions (all purpose modes)
8. **Calculate RCA Adjustments & Download** (appraisal only)
9. **Sales Comparison Grid** (appraisal only) --- generate formatted Excel
   workbook with comp selection modal
10. **Download Report** --- Word (.docx), PDF, or HTML

A **Skip first row** checkbox appears below the Purpose selector in
General and Market modes. Settings are saved per file and purpose
combination, so switching modes preserves each mode's configuration.
Changing the Purpose resets both the data import (Section 1) and the
earthUI import (Section 2) --- you must re-import your data after
switching purpose.

## Variable Selection

### The Predictor Table

Each column is listed with:

- **Type** --- data type dropdown (numeric, integer, character, Date, POSIXct)
- **Fac** --- factor checkbox. Check to treat the variable as a categorical
  factor (creates dummy columns in the model matrix)
- **Inc** --- include as predictor checkbox
- **Force** --- force into model (penalty factor = 0, never dropped by lasso)
- **Special** --- special column designation (appraisal/market only)
- **Sign** --- expected coefficient sign (positive, negative, or either)
- **NAs** --- missing value count

### Special Column Types

In appraisal and market modes, each predictor can be assigned a special type:

**Date & Time:**

- `contract_date` --- triggers `sale_age` computation
- `listing_date` --- fallback for DOM calculation
- `dom` --- Days on Market column

**Monetary:**

- `concessions` --- used in Net SP formula (Sale Price - Concessions)

**Size & Location:**

- `latitude` / `longitude` --- rounded to 3 dp
- `area` --- market area/neighborhood identifier (e.g., area_id, neighborhood)
- `living_area` --- enables per-SF residuals
- `lot_size` / `site_dimensions` --- site size columns

**Age:**

- `actual_age` / `effective_age` --- property age columns

**Display:**

- `display_only` --- in exports but excluded from model (multiple allowed)

**Weighting:**

- `weight` --- observation weight column (available in all modes; only one
  allowed; rows with weight = 0 are excluded from fitting)

### Expected Coefficient Signs

For numeric predictors, specify whether you expect the coefficient to be
**positive**, **negative**, or **either**:

- **Square footage**: Expected **positive** (more space = higher value)
- **Age**: Expected **negative** (older properties typically worth less)
- **Lot size**: Expected **positive**

If the model produces a coefficient with an unexpected sign, it is flagged
in the coefficient table.

### Enforce Sign Constraints

When the **Enforce Sign Constraints** checkbox is enabled, the model is
forced to produce coefficients matching the expected signs. A "positive"
sign forces the coefficient >= 0; "negative" forces <= 0; "either" is
unconstrained. This is implemented via `upper.limits` and `lower.limits`
in `glmnet`.

## Model Parameters

### Alpha (Mixing Parameter)

Controls the type of regularization:

- **0 (Ridge)**: Keeps all variables, shrinks coefficients proportionally.
- **1 (Lasso)**: Performs variable selection by setting some coefficients to
  zero. Default and recommended for appraisal work.
- **Between 0 and 1**: Elastic net blend.
- **Grid search**: Test multiple alpha values and select the best.

### Lambda Selection

Controls the strength of regularization:

- **Cross-validation (automatic)**: Recommended. Tests many lambda values
  using k-fold CV.
  - **lambda.1se**: More regularized (simpler model). Default.
  - **lambda.min**: Minimum cross-validation error.
- **Manual**: Enter a specific lambda value.

### Family

- **gaussian**: Continuous responses (e.g., sale price). Most common.
- **binomial**: Binary outcomes (e.g., sold/not sold).
- **poisson**: Count data.

### Relaxed Lasso

Regular lasso uses the same penalty for variable selection AND coefficient
estimation, which can over-shrink important coefficients. Relaxed lasso
separates these: first select variables via lasso, then refit with less or
no penalty.

- **Gamma = 0**: Fully relaxed (OLS refit, no shrinkage)
- **Gamma = 1**: No relaxation (same as regular glmnet)
- With cross-validation, gamma is chosen automatically

### Random Seed

A text input pre-filled with a random integer ensures reproducible
cross-validation fold assignments. After each fit, a new seed is
auto-generated (using system time, not R's RNG). The last 5 seeds used
for the current file are shown as clickable links for easy recall.
Seed history resets when a new file is loaded.

### Interaction Matrix

An interactive upper-triangular matrix lets you control which variable pairs
are allowed to interact. Check a cell to allow the interaction term
`x1:x2` in the model matrix.

- **Allow All**: Check all interaction pairs
- **Clear All**: Uncheck all pairs
- **Click a variable name**: Toggle all interactions for that variable
- Interaction selections persist via localStorage per input file

With many predictors, lasso will zero out unneeded interaction terms
automatically, so "Allow All" is a reasonable starting point.

> **Warning:** Enabling glmnet interactions will produce adjustments that
> can be difficult to interpret and explain in court or audit settings.
> Use interactions cautiously and verify that the resulting adjustments
> are reasonable.

### Block from Main Effect (Interaction Only)

**Right-click** a variable name in the Interaction Matrix to block it
from entering the model as a main effect. A bold **" 1"** indicator
appears after the variable name.

When blocked:

- The variable is **removed** from the main formula terms
- The variable **can still** appear in interaction terms (`x1:x2`)
- Right-click again to unblock
- State persists via localStorage per input file

**Use case:** When modeling time adjustments that should scale with
property size, block `sale_age` from the main effect and allow it only
in an interaction with `living_area`. This creates a `sale_age:living_area`
interaction term instead of a flat `sale_age` coefficient -- the time
appreciation rate varies by house size.

## Result Tabs

### Data Preview

Shows the imported data. In appraisal/market modes, split into Subject
Property (row 1) and Comparable Sales (rows 2+).

### Equation

LaTeX-rendered model equation via MathJax showing all non-zero coefficients.
Interaction terms are displayed as `x₁ × x₂`. Available after model fitting.

### Correlation

Heatmap matrix of Pearson correlations among numeric predictors and the
response variable. Available immediately after data import (no model fitting
required). Adaptive text and axis sizing based on variable count.

Color gradient: blue (-1) → white (0) → red (+1).

### Summary

Model fit statistics displayed as cards:

- **R²** --- proportion of variance explained
- **Adj R²** --- adjusted for number of predictors
- **GR²** --- generalized R² from glmnet's deviance ratio
- **CV R²** --- cross-validation R² (when CV is used)
- **RMSE** --- root mean squared error
- **MAE** --- mean absolute error

In appraisal/market modes, additional metrics:

- **Median Ratio** --- median of (predicted / actual)
- **COD** --- Coefficient of Dispersion (lower is better)
- **PRD** --- Price-Related Differential

An **overfitting warning** appears when training R² exceeds CV R² by >0.1.

### Coefficients

Table of all non-zero coefficients with sign violation warnings.

### Variable Importance

Bar chart showing standardized coefficient magnitude (|β| × sd(x)) per
predictor, aggregated across dummy columns for factor variables. Uses
interactive plotly (hover for exact values) when available, with ggplot2
fallback.

Importance table below shows: Variable, Importance, Coefficient, Relative %.

### Contributions

Per-variable partial effect plots, selected via dropdown:

- **Numeric predictors**: Scatter plot with linear fit line, labeled with
  the slope (e.g., `+1,234.56/unit`).
- **Factor variables**: Box plot showing contribution distribution per
  factor level.
- **Two-variable interactions**: Three stacked visualizations:
  1. 3D surface plot (`persp()`) of the joint contribution
  2. Scatter plot with points colored by contribution
  3. Heatmap of mean contribution across both variable ranges

### ANOVA

Variance decomposition table showing per-variable:

- Sum of Squares (SS)
- Percentage of model SS
- Coefficient

Includes intercept row and total model SS row.

### Diagnostics

Four diagnostic plots with large fonts, 15 axis tick marks, and
comma-formatted labels (no scientific notation):

1. **Coefficient Path**: Shows how coefficients change as lambda increases.
2. **CV Error**: Cross-validation error across lambda values with
   lambda.min and lambda.1se marked.
3. **Actual vs Predicted**: Points should cluster around the 45-degree line.
4. **Residuals vs Fitted**: Should show random scatter around zero.

Each plot has a **Download PNG** button.

### Glmnet Output

Raw model output showing: random seed used, model print, selected lambda,
lambda.min/1se, gamma (if relaxed), and full coefficient vector.

### Report

Configure report fields (appraiser name, property address, report date,
file number) and export to Word (.docx), PDF, or HTML. Reports are
rendered via Quarto when available, with an automatic rmarkdown fallback.

Report contribution plots match the Contributions tab: scatter with fit
line for numeric predictors, box plots for factors, and scatter/heatmap/3D
persp snapshots for interactions. All axis labels use comma-formatted
numbers (no scientific notation).

## Downloading Data (Step 7)

After fitting, download an Excel file with predictions and diagnostics.
In appraisal/market modes, the output includes:

| Column | Description |
|:-------|:------------|
| `est_<target>` | Model prediction (e.g., `est_sale_price`) |
| `residual` | Actual - predicted |
| `cqa` | Comparative Quality Analysis score (0-10 scale) |
| `residual_sf` | Residual / living area (if designated) |
| `cqa_sf` | CQA calculated from ranking via residual_sf |
| `<var>_contribution` | Per-variable contribution |
| `basis` | Intercept value (same for all rows) |

### CQA Scores

CQA ranks each row's residual against all others on a 0--10 scale:

- **High CQA (~9-10)**: sold for much more than predicted
- **Low CQA (~0-1)**: sold for much less than predicted
- **CQA ~5**: near the median

In appraisal/market modes, rows are sorted by `residual_sf` descending.

## RCA Calculations (Step 8, Appraisal Only)

### The RCA Dialog

- Choose **CQA** or **CQA per SF** score type
- Enter the subject's CQA score (0.00--9.99, default 5.00)
- Click **Calculate** to compute adjustments

### How It Works

1. Comparables' CQA scores and residuals are sorted
2. Linear interpolation maps your CQA value to a residual
3. Subject value = model prediction + interpolated residual

### Output Columns

| Column | Description |
|:-------|:------------|
| `subject_value` | Model prediction + interpolated residual |
| `<var>_adjustment` | Subject contribution - comp contribution |
| `residual_adjustment` | Subject residual - comp residual |
| `net_adjustments` | Sum of all adjustments |
| `gross_adjustments` | Sum of absolute adjustments |
| `adjusted_sale_price` | Comp sale price + net adjustments |

## Sales Comparison Grid (Step 9, Appraisal Only)

After computing RCA adjustments, a **Sales Grid** button appears. Clicking it
opens a modal dialog for comp selection.

### Comp Selection

Comps are split into two groups:

- **Recommended** (pre-checked): Gross adjustment < 25% of sale price, sorted
  by sale age (most recent first).
- **Other Comps**: Remaining comps, unchecked by default.

Select up to 30 comps. Click **Generate Sales Grid** to create the workbook.

### Workbook Layout

The generated Excel workbook contains up to 10 sheets (3 comps per sheet):

- **Header**: Subject and comp addresses, sale prices, and sale dates
- **Grouped rows**: Location (latitude, longitude, area), Site (lot size, site
  dimensions), and Age (actual age, effective age) rows with sub-items
- **Model variables**: One row per non-zero coefficient showing contributions
  and adjustment percentages
- **Residual features**: Unlocked cells for appraiser input on features not
  captured by the model
- **Summary rows**: Net Adjustment %, Gross Adjustment %, and Adjusted Sale
  Price with formulas

Sheets are protected but residual value cells remain editable for appraiser
input.

## Settings Persistence

glmnetUI saves settings to the browser's localStorage:

- **Per-file**: Variable selections, model parameters, interaction matrix
  (keyed by filename)
- **Global defaults**: Saved via "Save current as default" button
- **Locale**: Country and paper size preferences
- **Theme**: Light/dark mode toggle
- **Purpose mode**: The last-used purpose (General, For Appraisal, or
  Market Area Analysis) is remembered and restored on relaunch

When you re-import a previously used file, all settings are restored
automatically.

### Settings Options

- **Use last settings for input file** --- restore saved settings for this
  specific file
- **Use default settings** --- apply your saved global defaults
- **glmnet defaults** --- reset to factory defaults (alpha = 1, lambda = CV,
  lambda.1se, 10 folds, gaussian, standardize on)

## Differences from earthUI

glmnetUI and earthUI are companion tools for regularized regression. Key
differences:

| Feature | glmnetUI | earthUI |
|:--------|:---------|:--------|
| Method | Elastic net (glmnet) | MARS/Earth (earth) |
| Variable selection | Automatic via lasso penalty | Automatic via pruning |
| Interactions | Pairwise via interaction matrix | Up to degree 3 |
| Coefficient signs | Enforceable via constraints | Not directly enforceable |
| Hinge functions | No (linear effects only) | Yes (piecewise linear) |
| Relaxed fitting | Relaxed lasso | Not applicable |
| Alpha tuning | Grid search available | Not applicable |
| Sales Grid | Available (Step 9) | Available (Step 8) |
| g-Functions | Not applicable | Grouped term visualization |

Both tools share the same data format, special column types, RCA workflow,
and demo datasets.

## Demo Dataset: Appraisal_1.csv

The demo dataset is shared with earthUI. If earthUI is installed, load it
with:

```{r eval=FALSE}
demo_file <- system.file("extdata", "Appraisal_1.csv", package = "glmnetUI")
```

Or import it directly through the Shiny app file upload.

### Description

The file contains 1,502 residential sales (plus 1 subject property in row 1)
from a simulated MLS export. The data represents single-family home sales
in a multi-area market with a range of property sizes, ages, and locations.

### Key Columns

| Column | Type | Special Type | Description |
|:-------|:-----|:-------------|:------------|
| `weight` | numeric | weight | Observation weight (0 = exclude) |
| `sale_price` | numeric | (target) | Sale price --- response variable |
| `living_sqft` | numeric | `living_area` | Gross living area in square feet |
| `baths_total` | numeric | --- | Total bath count (e.g., 2.5) |
| `lot_size` | numeric | `lot_size` | Lot size in square feet |
| `area_id` | integer | `area` | MLS area identifier |
| `age` | numeric | `actual_age` | Property age in years |
| `latitude` | numeric | `latitude` | Latitude (rounded to 3 dp) |
| `longitude` | numeric | `longitude` | Longitude (rounded to 3 dp) |
| `garage_spaces` | integer | --- | Number of garage bays |
| `contract_date` | Date | `contract_date` | Sale contract date |
| `days_on_market` | integer | `dom` | Days on market |
| `sale_concessions` | numeric | `concessions` | Seller concessions |
| `street_address` | character | `display_only` | Property address |

### Suggested Quick Start

1. Launch glmnetUI: `glmnetUI::glmnetUI()`
2. Import `Appraisal_1.csv` via the file upload
3. Set Purpose to **For Appraisal**
4. Select `sale_price` as the target
5. Assign special types as shown in the table above
6. Include predictors: `sale_age`, `living_sqft`, `baths_total`, `lot_size`,
   `area_id` (as factor), `age`, `latitude`, `longitude`, `garage_spaces`
7. Keep alpha = 1 (Lasso), CV with lambda.1se
8. Click **Fit Glmnet Model**
9. Review Summary, Coefficients, and Diagnostics tabs
10. Download output (Step 7), review the CQA ranking
11. Compute RCA adjustments (Step 8) with a CQA score of ~5.00
12. Generate a Sales Comparison Grid (Step 9) with recommended comps
