---
title: "Getting Started with glmnetUI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with glmnetUI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Prerequisites

glmnetUI works on macOS, Windows, and Linux with R >= 4.1.0. All features
work out of the box except:

- **PDF reports** require a LaTeX installation. Install with
  `tinytex::install_tinytex()`. If LaTeX is not detected, the PDF
  option is automatically hidden and HTML/Word reports remain available.
- **Roboto Condensed font** is downloaded from Google Fonts for plot
  styling. If your machine is offline, the app uses the system
  sans-serif font automatically.

## Introduction

`glmnetUI` provides an interactive Shiny interface for elastic net regression
modeling using the `glmnet` package. It runs as a local Shiny application ---
no login, no server, no accounts. You launch it from R, import a dataset
(CSV or Excel), configure your model, and fit it interactively.

**Elastic net regression** combines two regularization techniques:

- **Ridge regression** (alpha = 0): Shrinks coefficients toward zero but keeps
  all variables.
- **Lasso regression** (alpha = 1): Sets some coefficients exactly to zero,
  performing automatic variable selection.
- **Elastic net** (0 < alpha < 1): A blend of both approaches.

### Three Purpose Modes

Every project has a **purpose**, chosen when you create it (Section 1,
*Project*):

- **General** --- Elastic net regression for any dataset.
  A **Skip first row** checkbox lets you exclude row 1 from fitting.
- **For Appraisal** --- Tailored for real estate appraisal with subject
  property handling, special column designations, RCA adjustments, and
  formatted report export.
- **Market Area Analysis** --- For analyzing groups of properties with optional
  first row exclusion.

The active project's purpose determines which tools and interface elements
appear; to work in a different mode, create or open a project with that
purpose. Settings are saved per file and purpose combination, so each
mode's configuration is preserved independently. Opening, closing, or
switching a project resets the data import (Section 2), the earthUI import
(Section 3), and all model results.

### Launching the Application

```{r}
library(glmnetUI)
glmnetUI()
```

This opens the application at `http://localhost:7879`.

## Basic Workflow

### Create or open a project

Work is organized into **projects** under a shared `regProj` folder. In
**Section 1 (Project)**, click **+ New…**, choose a Purpose and a location
(the administrative levels you fill in depend on the country, so foreign
conventions are supported), and type a short **Project Name** of at most
**8 characters**.

The app automatically builds the project's folder name by prepending the
location codes and the creation date-time:

```
<country>_<admin level codes…>_<YYYYMMDD-HHMMSS>_<your name>
```

e.g. `us_ca_081_burlin_20260604-131500_burl`. Only newly created projects get
this prefix; existing projects keep their names. See the User Guide for
details. Once a project is open, continue with the steps below.

### 2. Import Data

In **Section 2 (Import Data)**, pick a file from the active project's
`<os>_in/` folder (add files via Finder/Explorer or the **Browse...**
button, then click **Refresh**). Supported formats:

| Format | Extension | Notes |
|--------|-----------|-------|
| CSV | .csv | Locale-aware separator and decimal mark |
| Excel | .xls, .xlsx | Sheet selector for multi-sheet workbooks |

Column names are automatically converted to `snake_case` on import.

### 3. Import from earthUI (Optional)

Section 3 of the sidebar lets you import an earthUI result `.rds` file.
This allows glmnetUI to use earth's hinge basis functions (piecewise
linear terms) with glmnet's elastic net regularization. When imported,
the predictor set is dictated by the earth model: the variable table
locks the Include and Type columns, while Force and Sign stay editable.
This step is optional --- skip it if you do not have an earthUI result
to import.

### 4. Configure Variables

After importing, a predictor table shows each column with:

- **Type** --- data type dropdown (numeric, integer, character, Date, POSIXct)
- **Include** --- checkbox to include as predictor
- **Factor** --- factor checkbox (treat as categorical)
- **Force** --- force into model (penalty factor = 0, never dropped by lasso)
- **Sign** --- expected coefficient sign (positive, negative, or either)
- **NAs** --- missing value count
- **Special** --- special column designation (appraisal/market modes only)

Select your **response variable** from the dropdown, then check/uncheck
predictors as needed.

### 5. Set Model Parameters

Under **"5. glmnet Call Parameters"**:

- **Alpha**: fixed value (0 = Ridge, 1 = Lasso, the default) or grid search
- **Lambda**: Cross-validation (recommended) or manual entry
- **Lambda choice**: `lambda.1se` (simpler model) or `lambda.min` (best fit)
- **Family**: gaussian (continuous), binomial (binary), or poisson (counts)
- **Standardize**: Recommended on (default)
- **Enforce sign constraints**: Force coefficients to match expected signs
- **Relaxed lasso**: Refit surviving variables with less shrinkage
- **Random seed**: reproducible CV folds; the last 5 seeds are recallable
- **Interaction matrix**: Check pairs to allow interaction terms; all
  pairs start unchecked. Right-click a variable name to block its main
  effect while keeping it available in interactions.
  (Warning: glmnet interactions produce adjustments that can be hard to
  interpret in court or audit settings.)

### 6. Fit the Model

Click **"Fit Model"**. The application:

1. Prepares data (missing values, factor encoding, weights)
2. Runs cross-validation to find optimal lambda
3. Fits the final model
4. Populates all result tabs

### 7. Review Results

Results appear in the main panel tabs:

- **Data Preview** --- Raw data table (double-click a cell to see its full text)
- **Equation** --- LaTeX-rendered model equation
- **Correlation** --- Heatmap of predictor correlations (no model needed)
- **Summary** --- R², GR², CV R², RMSE, MAE; appraisal metrics (COD, PRD)
- **Coefficients** --- Non-zero coefficients with sign warnings
- **Variable Importance** --- Bar chart of |coef| x sd(x) per predictor
- **Contributions** --- Per-variable effect plots: scatter with fit line
  (numeric), box plot (factor), scatter/heatmap/3D surface (interactions)
- **ANOVA** --- Variance decomposition table
- **Diagnostics** --- Coefficient path, CV error, actual vs predicted, residuals
- **Glmnet Output** --- Raw model output, selected lambda, and the random
  seed used
- **Report** --- Configure report fields and export to Word or PDF
- **RCA Adjustments** --- RCA plots and tables (appraisal/market, after
  Step 8)

### 8. Export

- **Download Output** (all modes): Excel with predictions, residuals, CQA
  scores, and per-variable contributions
- **Calculate RCA Adjustments & Download** (appraisal and market modes):
  per-comparable adjustments with the subject residual interpolated from
  your CQA score
- **Generate Sales Grid & Download** (appraisal mode): Generate a formatted
  Sales Comparison Grid Excel workbook from RCA output. A modal dialog lets
  you select comps (recommended comps with gross adjustment < 25% are
  pre-checked). Up to 30 comps across 10 sheets (3 per sheet) with grouped
  rows, formulas, and unlocked residual cells for appraiser input.
- **Quarto reports** (two steps): **Generate Quarto Report** writes a
  self-contained `.qmd` bundle with all tables and plots to the project's
  output folder; **Convert Quarto Report** renders it to HTML, Word, or
  PDF. Reports include contribution plots matching the Contributions tab
  (scatter, heatmap, 3D persp snapshots for interactions). Rendered via
  Quarto with automatic rmarkdown fallback. The Report tab also offers
  direct Word/PDF export.

## Tips

1. **Start with Lasso (alpha = 1)** to identify important variables.
2. **Use lambda.1se** for simpler, more defensible models.
3. **Set expected signs** before fitting and review warnings after.
4. **Use interactions cautiously** -- glmnet interactions can produce adjustments
   that are hard to interpret for court or audit purposes.
5. **Try relaxed lasso** if coefficients appear over-shrunk.

## Demo Dataset

glmnetUI includes the `Appraisal_1.csv` demo dataset:

```{r}
demo_file <- system.file("extdata", "Appraisal_1.csv", package = "glmnetUI")
```

To use it in the app, copy it into your project's `<os>_in/` folder (or
select it via the **Browse...** button in Section 2).

See the **User Guide** vignette for complete documentation of all features,
special column types, RCA calculations, and the demo dataset.
